A tutorial on Bayesian and Frequentist Event History Analyses for psychological time-to-event data

### 2024-08-07 manuscript/manuscript.Rmd
update manuscript

```{r setup, include = FALSE}
pkg <- c("papaja", "citr", "tidyverse", "RColorBrewer", "patchwork")

lapply(pkg, library, character.only = TRUE)

r_refs(file = "r-references.bib")
my_r_citation <- cite_r(file = "r-references.bib", footnote = TRUE)
```

@@

Similarly, after dividing time in discrete, contiguous time bins indexed by t, let RT be a discrete random variable denoting the rank of the time bin in which a particular person's response occurs in a particular experimental condition. Discrete-time EHA focuses on the discrete-time survivor function S(t) = P(RT $>$ t) and the discrete-time hazard function h(t) = P(RT = t| RT $\geq$ t), and not on the probability mass function and the cumulative distribution function. The discrete-time hazard probability function gives you the probability that the event occurs (sometime) in bin t, given that the event has not occurred yet in previous bins. For two-choice RT data, the discrete-time hazard function can be extended with the conditional accuracy function ca(t) = P(correct | RT = t), which gives you the probability that a response is correct given that it has been emitted in time bin t [@kantowitzInterpretationReactionTime2021; @wickelgrenSpeedaccuracyTradeoffInformation1977; @allisonSurvivalAnalysisUsing2010]. This latter function is also known as the micro-level speed-accuracy tradeoff function.

@@

We used `r my_r_citation$r` for all reported analyses. Web links are printed in bold.

@@

In this second tutorial we illustrate how to fit a Bayesian hazard regression model for the data set used in the first tutorial.

model formula 
analytic decisions: specification of the effect of TIME,

assumptions:
- linearity: continuous variables (TIME, trial number) 
- additivity:
- proportionality:

>>>>

```{r setup, include = FALSE}
# Tutorial 1 and 4
pkg1 <- c("papaja", "citr", "tidyverse", "RColorBrewer", "patchwork")

lapply(pkg1, library, character.only = TRUE)

# Tutorial 2
pkg2 <- c("cmdstanr", "standist", "tidyverse", "RColorBrewer", "patchwork", 
         "brms", "tidybayes", "bayesplot", "future", "parallel")

lapply(pkg2, library, character.only = TRUE)

options(brms.backend = "cmdstanr",
        mc.cores = parallel::detectCores(),
        future.fork.enable = TRUE,
        future.rng.onMisuse = "ignore") ## automatically set in RStudio

supportsMulticore()
detectCores()

# Tutorial 3

# R references
r_refs(file = "r-references.bib")
my_r_citation <- cite_r(file = "r-references.bib", footnote = TRUE)
```

@@

Similarly, after dividing time in discrete, contiguous time bins indexed by t, let RT be a discrete random variable denoting the rank of the time bin in which a particular person's response occurs in a particular experimental condition. Discrete-time EHA focuses on the discrete-time hazard function h(t) = P(RT = t| RT $\geq$ t) and the discrete-time survivor function S(t) = P(RT $>$ t) = [1-h(t)].[1-h(t-1)].[1-h(t-2)]...[1-h(1)], and not on the probability mass function p(t) = h(t).S(t-1) and the cumulative distribution function F(t) = 1-S(t). The discrete-time hazard probability function gives you the probability that the event occurs (sometime) in bin t, given that the event has not occurred yet in previous bins. Unlike the hazard function, which assesses the unique risk associated with each time bin, the survivor function cumulates the bin-by-bin risks of event *non*occurrence.
For two-choice RT data, the discrete-time hazard function can be extended with the conditional accuracy function ca(t) = P(correct | RT = t), which gives you the probability that a response is correct given that it has been emitted in time bin t [@kantowitzInterpretationReactionTime2021; @wickelgrenSpeedaccuracyTradeoffInformation1977; @allisonSurvivalAnalysisUsing2010]. This latter function is also known as the micro-level speed-accuracy tradeoff function.

@@

We used `r my_r_citation$r` for all reported analyses. Web links are printed in bold. The content of this tutorial is mainly based on @allisonSurvivalAnalysisUsing2010, @singerAppliedLongitudinalData2003, @mcelreathStatisticalRethinkingBayesian2018, @kurzAppliedLongitudinalDataAnalysis2023, and @kurzStatisticalRethinkingSecondEd2023.

@@

When you want to study how hazard depends on various predictors, you can fit regression models to the data [@singerAppliedLongitudinalData2003]. There are two analytic decisions one has to make. First, one has to select an analysis time range, i.e., a contiguous set of bins for which there is enough data for each participant. Second, one can choose the logit link function which transforms a (hazard) probability into the log of the odds ratio, or the complementary log-log (cloglog) link function, which yields the logarithm of the negated logarithm of the probability of event *non*occurrence. An important difference between these two link functions is that cloglog provides a discrete-time hazard model that has a built-in proportional hazards assumption, while logit provides a proportional odds assumption (see below). The cloglog link is preferred over the logit link when events can occur in principle at any time point within a bin, which is the case for RT data [@singerAppliedLongitudinalData2003]. Third, one has to choose a specification of the effect of discrete TIME (i.e., the time bin index t). One can choose a general specification (one intercept per bin) or a functional specification, such as a polynomial one.

An example discrete-time hazard model with three predictors (TIME, X1, X2) and the cloglog link function can be written as follows:

cloglog[h(t)] = ln(-ln[1-h(t)]) =  [$\alpha$~1~ONE + $\alpha$~2~(TIME-1) + $\alpha$~3~(TIME-1)$^2$] +                                    [$\beta$~1~X~1~ + $\beta$~2~X~2~ + $\beta$~3~X~2~(TIME-1)].

The main predictor variable TIME is the time bin index t that is centered on value 1 in this example. The first set of terms within brackets, the alpha parameters multiplied by their polynomial specifications of (centered) time, represents the shape of the baseline cloglog-hazard function (i.e., when all predictors X~i~ take on a value of zero). The second set of terms (the beta parameters) represents the vertical shift in the baseline cloglog-hazard for a 1 unit increase in the respective predictor. Predictors can be discrete, continuous, and time-varying or time-invariant. For example, the effect of a 1 unit increase in X~1~ is to vertically shift the whole baseline cloglog-hazard function by $\beta$~1~ cloglog-hazard units. However, if the predictor interacts linearly with time (see X~2~ in the example), then the effect of a 1 unit increase in X~2~ is to vertically shift the predicted cloglog-hazard in bin 1 by $\beta$~2~ cloglog-hazard units (when TIME-1 = 0), in bin 2 by $\beta$~2~ + $\beta$~3~ cloglog-hazard units (when TIME-1 = 1), and so forth. To interpret the effects of the predictors, the parameter estimates are exponentiated, resulting in a hazard ratio (due to the use of the cloglog link).

In the case of a large-*N* design without repeated measurements, the parameters of a discrete-time hazard model can be estimated using standard logistic regression software (after expanding the typical person-trial-oriented data set into a person-trial-bin-oriented data set (see Tutorial 1); @allisonSurvivalAnalysisUsing2010). When there is clustering in the data, as in the case of a small-*N* design with repeated measurements, the parameters of a discrete-time hazard model can be estimated using population-averaged methods (e.g., Generalized Estimating Equations), and Bayesian or frequentist generalized linear mixed models [@allisonSurvivalAnalysisUsing2010].

In this second tutorial we illustrate how to fit a Bayesian hazard regression model for the masked response priming data set used in the first tutorial. 
In general, there are three assumptions one can make or relax when adding experimental predictor variables: The linearity assumption for continuous predictors (the effect of a 1 unit change is the same anywhere on the scale), the additivity assumption (predictors do not interact), and the proportionality assumption (predictors do not interact with TIME).

First, we select the analysis range (200,600] and the cloglog link, and use a polynomial to specify the effect of TIME in the "blank" prime condition. Second, based on previous work [@panisHowCanWe2020; @panisNeuropsychologicalEvidenceTemporal2017; @panisStudyingDynamicsVisual2020; @panisTimecourseContingenciesPerceptual2009; @panisWhenDoesInhibition2022] and because cognition is likely the behavior of a non-linear dynamical system [ref], we relax all three assumptions, as follows:

\scriptsize
```{r prepare-data, echo=T}
# load data
ptb_data <- read_csv("../Tutorial_1_descriptive_stats/data/inputfile_hazard_modeling.csv")
# select analysis time range: (200,600] with 10 bins (time bin ranks 6 to 15)
ptb_data <- ptb_data %>% filter(period > 5)
# create factor condition, with "blank" as the reference level
ptb_data <- ptb_data %>% mutate(condition = factor(condition, labels = c("blank", "congruent","incongruent")))
# center discrete TIME (period) on bin 9, and trial on trial 1000
ptb_data <- ptb_data %>% mutate(period_9 = period - 9,
                                trial_c = (trial - 1000)/1000)
# remove unnecessary columns before fitting a model
ptb_data <- ptb_data %>% select(-c(bl,tr,trial,period)) # 12840 obs. of 5 variables
```

```{r set-priors, echo=T}
priors <- c(
  set_prior("normal(0, 1)", class = "b"), # for beta parameters 
  set_prior("student_t(7.61, 0, 1.57)", class = "b", coef = "Intercept"), # flat prior for intercept on hazard scale
  set_prior("normal(0, 1)", class = "sd"), # for standard deviation of RE
  set_prior("lkj(2)", class = "cor") # for correlations between RE
)
```

```{r fit-model, echo=T}
plan(multicore)
model_full_RE <-
   brm(data = ptb_data,
       family = binomial(link="cloglog"),
       event | trials(1) ~ 0 + Intercept + 
                           condition*period_9*trial_c + 
                           condition*I(period_9^2) + 
                           condition*I(period_9^3) +
                           (1 + condition*period_9*trial_c +
                           condition*I(period_9^2) +
                           condition*I(period_9^3) | pid),
       prior = priors,
       chains = 4, cores = 4, iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, step_size = 0.04, max_treedepth = 12),
       seed = 12, init = "0",
       file = "../Tutorial_2_Bayesian/models/model_full_RE")
```
\normalsize

To test whether (centered) trial number affects behavior, we fit a model without the variable trial_c.

Use WAIC to compare models.

Plot the effects of congruent and incongruent for each time bin for the selected model.

Plot the model-based hazard and survivor functions.
